---
title: "Lab 09: Algorithmic Bias"
author: "Haley Lam"
date: "3/15/26"
output: github_document
---

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message = FALSE}
library(tidyverse)
library(fairness)
library(janitor)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.


Clean the code (given in the textbook online)
```{r}
# Load the COMPAS data
compas <- read_csv("data/compas-scores-2-years.csv") %>%
  clean_names() %>%
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )

# Take a look at the data
glimpse(compas)
```

## Part 1: Exploring the Data

### Exercise #1
Each row represents an individual, and the columns have variables about that individual (e.g., sex, age, race)
There are `r nrow(compas)` individuals and `r ncol(compas)` pieces of information about each individual.

### Exercise #2
```{r}
# number of duplicates in the dataset (based on name)
sum(duplicated(compas$name))
```
There are `r sum(duplicated(compas$name))` duplicates! They might be defendants who were arrested multiple times?

### Exercise #3
Let's examine the distribution of decile_score
```{r}
ggplot(compas, aes(x = decile_score)) +
  geom_bar()
```
It is not normally distributed and right-skewed! Most people are given a low decile_score.

### Exercise #4
Distribution of demographics in the dataset.
```{r}
# Sex
ggplot(compas, aes(x = sex))+
  geom_bar()

# Race
ggplot(compas, aes(x = race))+
  geom_bar()

# Age
ggplot(compas, aes(x = age_cat))+
  geom_bar()
```

## Part 2: Risk Scores and Recidivism

### Exercise #5
What's the relationship between risk scores and recidivism?
```{r}
compas %>%
  group_by(decile_score) %>%
  summarize(rate = mean(two_year_recid, na.rm = TRUE)) %>%
  ggplot(aes(x = decile_score, y = rate)) +
  geom_point()
```
Wow! That is a very positively correlated scatterplot if I've seen one! 

### Exercise #6
The "accuracy" of the algorithm
```{r}
compas <- compas %>%
  mutate(compas_classification = case_when(
    decile_score >= 7 & two_year_recid == 1 ~ "TP",
    decile_score <= 4 & two_year_recid == 0 ~ "TN",
    decile_score >= 7 & two_year_recid == 0 ~ "FP",
    decile_score <= 4 & two_year_recid == 1 ~ "FN",
    TRUE ~ NA_character_
  ))

compas %>% 
  count(compas_classification)
```
There are 1,216 false negatives and 644 false positives. There are 2,681 true negatives and 1,351 true positives. 

### Exercise #7
```{r}
compas %>%
  filter(!is.na(compas_classification)) %>%
  summarize(
    accuracy = sum(compas_classification %in% c("TP", "TN")) / n()
  )
```
Filtering out the NA values, the model got 68.4% correct. This isn't great!



