---
title: "Lab 09: Algorithmic Bias"
author: "Haley Lam"
date: "3/15/26"
output: github_document
---

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message = FALSE}
library(tidyverse)
library(fairness)
library(janitor)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.


Clean the code (given in the textbook online)

```{r}
# Load the COMPAS data
compas <- read_csv("data/compas-scores-2-years.csv") %>%
  clean_names() %>%
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )

# Take a look at the data
glimpse(compas)
```

## Part 1: Exploring the Data

### Exercise #1
Each row represents an individual, and the columns have variables about that individual (e.g., sex, age, race)
There are `r nrow(compas)` individuals and `r ncol(compas)` pieces of information about each individual.

### Exercise #2

```{r}
# number of duplicates in the dataset (based on name)
sum(duplicated(compas$name))
```
There are `r sum(duplicated(compas$name))` duplicates! They might be defendants who were arrested multiple times?

### Exercise #3
Let's examine the distribution of decile_score

```{r}
ggplot(compas, aes(x = decile_score)) +
  geom_bar()
```

It is not normally distributed and right-skewed! Most people are given a low decile_score.

### Exercise #4
Distribution of demographics in the dataset.

```{r}
# Sex
ggplot(compas, aes(x = sex))+
  geom_bar()

# Race
ggplot(compas, aes(x = race))+
  geom_bar()

# Age
ggplot(compas, aes(x = age_cat))+
  geom_bar()
```

## Part 2: Risk Scores and Recidivism

### Exercise #5

What's the relationship between risk scores and recidivism?

```{r}
compas %>%
  group_by(decile_score) %>%
  summarize(rate = mean(two_year_recid, na.rm = TRUE)) %>%
  ggplot(aes(x = decile_score, y = rate)) +
  geom_point()
```

Wow! That is a very positively correlated scatterplot if I've seen one! 

### Exercise #6

The "accuracy" of the algorithm

```{r}
compas <- compas %>%
  mutate(compas_classification = case_when(
    decile_score >= 7 & two_year_recid == 1 ~ "TP",
    decile_score <= 4 & two_year_recid == 0 ~ "TN",
    decile_score >= 7 & two_year_recid == 0 ~ "FP",
    decile_score <= 4 & two_year_recid == 1 ~ "FN",
    TRUE ~ NA_character_
  ))

compas %>% 
  count(compas_classification)
```

There are 1,216 false negatives and 644 false positives. There are 2,681 true negatives and 1,351 true positives. 

### Exercise #7

```{r}
compas %>%
  filter(!is.na(compas_classification)) %>%
  summarize(
    accuracy = sum(compas_classification %in% c("TP", "TN")) / n()
  )
```
Filtering out the NA values, the model got 68.4% correct. This isn't great!

## Part 3: Investigating Disparities
### Exercise #8

```{r}
compas %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  ggplot(aes(x = decile_score)) +
  geom_bar() +
  facet_wrap(~ race)
```
The Caucasian one is right-skewed (like the overall distribution), while the African-American one is evenly distributed (like a uniform distribution).

### Exercise #9

```{r}
compas %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race) %>%
  summarize(pct_high_risk = mean(decile_score >= 7, na.rm = TRUE))
```

36.8% of Black defendants were classified as high risk, while 17.1% of Caucasian defendants were classified as high risk.

### Exercise #10
 Proportion of non-recidivists who were classified as high risk

```{r}
non_recidivists <- compas %>%
  filter(two_year_recid == 0)

fpr <- non_recidivists %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race) %>%
  summarize(rate = mean(decile_score >= 7, na.rm = TRUE)) %>%
  mutate(error_type = "False Positive Rate")
fpr
```

There's a 25% false-positive rate for African-Americans while a 9% false-positive rate for Caucasians.
 
 
Proportion of recidivists who were classified as low risk

```{r}
recidivists <- compas %>%
  filter(two_year_recid == 1)

fnr <- recidivists %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race) %>%
  summarize(rate = mean(decile_score <= 4, na.rm = TRUE)) %>%
  mutate(error_type = "False Negative Rate")
fnr
```

There's a 28% false-negative rate for African-Americans and a 48% false-negative rate for Caucasians.

### Exercise #11

Create a plot to visualize the discrepency

```{r}
bind_rows(fpr, fnr) %>%
  ggplot(aes(x = race, y = rate)) +
  geom_col() + 
  facet_wrap(~ error_type)
```

The false negative bar is higher for Caucasians, and false positive bar is higher for African-Americans.

## Part 4: Understanding Sources of Bias
### Exercise #12
```{r}
compas %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  ggplot(aes(x = priors_count, y = decile_score, color = race)) +
  geom_point() +
  geom_smooth(method = "lm")
```
It seems the slope for Caucasian is greater than the slope for African-American, though African-Americans have a higher score with 0 priors. This means the algorithm assigns a higher risk score to African-Americans even when they have no prior convictions.

### Exercise #13
Check calibration:
```{r}
compas %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race, decile_score) %>%
  summarize(rate = mean(two_year_recid, na.rm = TRUE)) %>%
  ggplot(aes(x = decile_score, y = rate, color = race)) +
  geom_line()
```
It seems like the lines overlap pretty well, showing good calibration. This supports Northpointe’s claim that the algorithm is fair!

## Part 5: Designing Fairer Algorithms
### Exercise #14
How do we create a fairer risk assessment algorithm?

I assume that variables like neighborhood (and its crime rates), zip code, etc. were collected and used for the algorithm. These variables might highly correlate with race and socioeconomic status. If they were removed, it's possible that there will be less bias, even though model "accuracy" might decrease. 

Another idea is to try adjust the model to make the 'false positive rate' roughly the same across racial groups. There'll be a tradeoff between accuracy and this type of 'fariness', but aiming for this equality in false positive rates seems fair even if overall accuracy drops.

### Exercise #15
ProPublica's conceptualization of fairness is about equal error rates across races. Northpointe's is about the algorithm's score being equally predictive of risk across races. There is a tradeoff between the model's accuracy (i.e., how well the model can predict risk overall) and how equal it is for races (i.e., how much error discrepency there is across races). 

### Exercise #16
It seems necessary to me for these tools to have openly accessible data for the public to know what these algorithms are trained on. It is also important that the false positive or false negative rates for each race (or other variables) are known to judges. They can make their decisions more clearly with this information and can choose to trust the computer less if a specific race has unequal error rates.


